Highly Available Transactions: Virtues and Limitations

### 2013-04-28 writeup/hat-analysis.tex
more examples, writing tweaks
some graphs and joe comments

To begin, we present well-known semantics that can be achieved in HAT
systems. We offer proof-of-concept highly available algorithms, but
our goal is \textit{not} to provide optimal or even efficient
implementations. As is standard in the distributed database
literature, in our sample implementations, we consider a set of
servers acting as replicas for a set of data items with which
non-co-located clients interact. In our examples, we exclusively
consider read and write operations, denoting a write of value $v$ to
data item $d$ as $w_d(v)$ and a read from data item $d$ returning $v$
as $r_d(v)$. We assume that all data items have the null value,
$\bot$, at database initialization, and, unless otherwise specified,
all transactions in the examples commit.

@@

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}, requiring that each transaction's writes are ordered
consistently according to a single total order on transactions. This
prohibits Adya's phenomenon $G0$, also called ``Dirty
Writes''~\cite{adya}. More formally, we require a total ordering on
writes to each data item and we create a graph of transactions with
edges from one transaction to another if the former overwrites the
latter's write to the same object. \textit{PL-1} prohibits cycles in
this graph, requiring some total order on writes. Accordingly, the
final database state cannot contain the ``earlier'' transaction's
writes to the data items: for example, in the below example, $T_3$
should \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2,
b=1$ or $a=1, b=2$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
\end{align*}
Later properties will strengthen Read Uncommitted.

Unlike the total order required by serializability, the order in PL-1
does not constrain the values seen in by a transaction's read
operations except once the database has reached a ``final
state.'' Read Uncommitted is easily achieved via applying the same
logical timestamp to each update in a transaction and applying a
``last writer wins'' conflict reconciliation policy at each replica.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs. Implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. The implementations often provide recency and monotonicity
properties beyond the simple meaning from the name, which is what is
expressed in the implementation-agnostic definition: under Read
Committed, transactions should do not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's \textit{PL-2} and is formalized by prohibiting
Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
from Berenson et al.). For instance, in the example below, $T_3$
should never see $a=1$, and, if $T_2$ aborts, $T_3$ should never read
$a=3$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-.5em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of the databaseq, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. This implementation does not provide recency or
monotonicity guarantees but satisfies the implementation-agnostic
definition.

``Repeatable Read'' isolation is a confusing property. For example,
some IBM products use the term for fully serializable
isolation~\cite{hat-hotos}, while Gray~\cite{gray-isolation}, Berenson
et al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except for predicate-based reads (and so Repeatable Read is
identical to serializability in a key-value store without
predicate-based access). In this latter model, the \textit{Phantom
  Problem}, whereby two successive predicate-based reads return
different data~\cite{gray-isolation}, is the only kind of
non-serializable behavior allowed. In this paper, we reserve the use
of ``Repeatable Read'' to refer to these definitions, which we will
soon show are not HAT-compliant.

However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
specification allows several additional behaviors under ``Repeatable
Read'' isolation.  \textbf{ANSI Repeatable Read} requires that, along
with respecting Read Committed isolation, each transaction will only
read one version of each data item that it did not itself produce
(preventing ``Fuzzy Read,'' or P2). In the example below, $T_3$ must
read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This is a literal interpretation of the phrase ``Repeatable Read'':
unless a transaction modifies a given data item, the observable value
of the data item should not change during the transaction. By itself,
this property is weak---we can always read $\bot$ for each data
item---and easily achieved in a HAT system by caching the values
read. To prevent confusion, we call this isolation model \textbf{Item
  Cut Isolation}: all the transaction's reads should see values from a
non-changing consistent cut or snapshot over the data items. It is
possible to satisfy Item Cut Isolation with high availability by using
multi-versioning on each replica and ensuring that each of a
transaction's reads return an appropriately chosen version (e.g.,
identified by vector clock).

A stronger achievable property that expands on Item Cut Isolation
requires consistent cuts to span both a transaction's predicate reads
(e.g., \texttt{SELECT WHERE}) as well as its item reads.  We call this
\textbf{Predicate Cut Isolation} and it prohibits Phantoms. Predicate
Cut Isolation is also achievable in HAT systems via similar
multi-versioning as above.

@@

We can achieve TA with a two-phase multi-versioning algorithm. Clients
buffer their writes until commit, sending the final write to each data
item to each respective server. Along with each write, clients attach
a the transaction ID and a list of all data items written to in the
transaction, $L_v$. Upon receiving a new write to item $d$, a server
places the write into a \texttt{pending} queue and notifies all other
servers responsible for a data item in $L_v$ that it has received a
write to $d$ with the respective transaction ID. Once a server
receives acknowledgments with appropriate transaction ID for all items
in $L_v$, it moves the write from \texttt{pending} to the
\texttt{stable} database (which applies the writes according to
highest-ID-wins). Accordingly, all writes that should be made atomic
with a write in \texttt{stable} are guaranteed to be present on their
respective servers, either in \texttt{pending} or
\texttt{stable}. Given the possibility of internal aborts, writes to
\texttt{pending} may require 2 RTTs: one to check constraints and one
to execute the above (but these checks \textit{cannot} cause other
transactions to block, as would be the case in a two-phase
\textit{commit} protocol).

For reads, each client tracks which values it should read. It
maintains a mapping (vector) from data items to transaction IDs. Upon
transaction start, the client clears its mapping, and, when it
performs a read from a data item, it sends the server its current ID
for the item. If the client does not supply an ID, the server returns
the latest value from \texttt{stable}. If the ID is non-empty, the
server either returns the matching value from \texttt{pending} or a
value with higher ID from \texttt{pending}---along with $L_v$ for the
write. When the client receives a response, it updates its local
vector with $L_v$. Accordingly, even if transaction's writes to two
keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
and the replica for $y$'s \texttt{pending}, a client's reads from $x$
then $y$ will return a transactionally atomic pair of values because
the replica for $y$ will know to check \texttt{pending}.

This implementation is entirely masterless and both reads and writes
do not block for coordination. One key property is that we do not, in
general, dictate when writes become visible. Accordingly, it is highly
available. However, we can optimize write visibility delay by ensuring
that clients are sticky with disjoint groups of servers. Once all
transaction values are \texttt{pending} within a group of servers,
values can be made \texttt{stable} (as opposed to once they are
present on all servers). For example, in a multi-datacenter setting,
all clients and servers within a datacenter may form a sticky group.

@@

Our models have not yet considered interactions \textit{across}
transactions, other than that there exists some arbitrary ordering on
them (via Read Uncommitted). In the distributed systems literature,
many useful \textit{safety} guarantees span multiple,
non-transactional operations. In particular, \textit{session
  guarantees} are used to describe guarantees across transactions
within a given \textit{session}, ``an abstraction for the sequence
of...operations performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

@@

Berenson et al. define \textit{Lost Update} as when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider a database containing only the following
transactions:
\begin{align*}
\small\vspace{-.5em}
T_1 &: r_x(a) w_x(a+2)
\\T_2 &: w_x(2)\vspace{-.5em}
\end{align*}
If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
operation, then the database will end up with $a=3$, a state that
could not have resulted in a serial execution due to from $T_2$'s
``Lost Update.''

@@

Their need to prevent Lost Update means that Consistent Read, Snapshot
Isolation, and Cursor Stability guarantees are all unavailable.
Repeatable Read (defined by to Gray, Berenson et al., and Adya) and
One-Copy Serializability need to prevent both Lost Update and Write
Skew. These prevention requirements mean that these models are
inherently unavailable.

@@

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Interestingly, strong one-copy serializability subsumes all
other models, while considering the (large) power set of all
compatible models (e.g., the diagram depicts 96 possible HAT
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonomy is not exhaustive, but we believe it
lends substantial clarity into the relationships between a large
subset of the prominent ACID and distributed consistency
models. Additional read/write transaction semantics that we have
omitted should be easily classifiable based on the available
primitives and HAT-incompatible anomaly prevention we have already
discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Linearizability\linearizable, Strong
1SR\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

\end{tikzpicture}
\label{fig:hat-order}
\caption{HAT, sticky HAT (in boxes), and unavailable models (circled)
  from Table~\protect\ref{table:hatcompared} compared
  graphically. Directed edges represent ordering by model
  strength. Models that do not share a common ancestor can be
  simultaneously achieved, and the resulting availability is that of
  the weakest available model in the combination.}
\label{fig:hatcompared}
\end{figure}

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. Transactional atomicity
dictates that writes to multiple keys across multiple servers are made
visible to readers all at once, while linearizability dictates that
writes to a single key on multiple servers are made visible to all
readers at once---what is different? First, in linearizable (and safe
and regular) systems, writes are made visible to clients
\textit{immediately} after they finish. With transactional atomicity,
there is no recency guarantee. Second, in linearizabile systems, all
clients see all writes at the same time. With transactional atomicity,
clients may see writes at different times depending on which replicas
they contact. We are not aware of an analogous model in the
distributed systems literature. Accordingly, despite apparent
similarities, transactional atomicity is incomparable with and much
cheaper (by availability standards) than linearizability.

>>>>

To begin, we present well-known semantics that can be achieved in HAT
systems. In this section, our goal is feasibility, not performance. As
a result, we offer proof-of-concept highly available algorithms, but
our goal is \textit{not} to provide optimal or even efficient
implementations: the challenge is to provide algorithms that provide
high availability. However, we briefly study performance implications
in Section~\ref{sec:evaluation}.

In our examples, we exclusively consider read and write operations,
denoting a write of value $v$ to data item $d$ as $w_d(v)$ and a read
from data item $d$ returning $v$ as $r_d(v)$. We assume that all data
items have the null value, $\bot$, at database initialization, and,
unless otherwise specified, all transactions in the examples commit.

@@

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}, requiring only that each transaction's writes are ordered
consistently according to a single total order on transactions. This
prohibits Adya's phenomenon $G0$, also called ``Dirty
Writes''~\cite{adya}. More formally, we require a total ordering on
writes to each data item and we create a graph of transactions with
edges from one transaction to another if the former overwrites the
latter's write to the same object. \textit{PL-1} prohibits cycles in
this graph, requiring some total order on writes. Accordingly, the
final database state cannot contain the ``earlier'' transaction's
writes to the data items: for example, in the below example, $T_3$
should \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2,
b=1$ or $a=1, b=2$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\vspace{-.5em}
\end{align*}
Later properties will strengthen Read Uncommitted.

Unlike the total order required by serializability, the order in PL-1
does not constrain the values seen by a transaction's read
operations except once the database has reached a ``final
state.'' Read Uncommitted is easily achieved via applying the same
logical timestamp to each update in a transaction and applying a
``last writer wins'' conflict reconciliation policy at each replica.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs. Centralized
implementations differ, with some based on long-duration exclusive
locks and short-duration read locks~\cite{gray-isolation} and others
based on multiple versions. The implementations often provide recency
and monotonicity properties beyond the simple meaning from the name,
which is what is expressed in the implementation-agnostic definition:
under Read Committed, transactions should not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's \textit{PL-2} and is formalized by prohibiting
Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
from Berenson et al.). For instance, in the example below, $T_3$
should never see $a=1$, and, if $T_2$ aborts, $T_3$ should never read
$a=3$:
\begin{align*}
\small\vspace{-.5em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-.5em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of the database, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. This implementation does not provide recency or
monotonicity guarantees but satisfies the implementation-agnostic
definition.

\textbf{Repeatable Read} isolation is a confusing property. For example,
some IBM products use the term for fully serializable
isolation~\cite{hat-hotos}, while Gray~\cite{gray-isolation}, Berenson
et al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except for predicate-based reads (and so Repeatable Read is
identical to serializability in a key-value store without
predicate-based access). In this latter model, the \textit{Phantom
  Problem}, whereby two successive predicate-based reads return
different data~\cite{gray-isolation}, is the only kind of
non-serializable behavior allowed. In this paper, we reserve the use
of ``Repeatable Read'' to refer to these definitions, which we will
soon show are not HAT-compliant.

However, as Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
specification allows several additional behaviors under ``Repeatable
Read'' isolation.  \textbf{ANSI Repeatable Read} requires that, along
with respecting Read Committed isolation, each transaction will only
read one version of each data item that it did not itself produce
(preventing ``Fuzzy Read,'' or P2). In the example below, $T_3$ must
read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This is a literal interpretation of the phrase ``Repeatable Read'':
unless a transaction modifies a given data item, the observable value
of the data item should not change during the transaction. By itself,
this property is weak---we can always read $\bot$ for each data
item---and easily achieved in a HAT system by caching the values
read. To prevent confusion, we call this isolation model \textbf{Item
  Cut Isolation}: all the transaction's reads should see values from a
non-changing consistent cut or snapshot over the data items. It is
possible to satisfy Item Cut Isolation with high availability by using
multi-versioning on each replica and ensuring that each of a
transaction's reads return an appropriately chosen version (e.g.,
identified by vector clock).

A stronger achievable property that expands on Item Cut Isolation
requires consistent cuts to span both a transaction's predicate reads
(e.g., \texttt{SELECT WHERE}) as well as its item reads.  We call this
\textbf{Predicate Cut Isolation} and it prohibits Phantoms (but does
not guarantee serializability of item-based reads and
writes). Predicate Cut Isolation is also achievable in HAT systems via
similar multi-versioning as above.

@@

TA can be achieved via multi-versioning. If clients attach vector
clocks to every write (incrementing their own position for each
transaction) and replicas store all writes via multi-versioning, then
clients can safely determine which sets of data items to read; this
approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
bolt-on causal consistency~\cite{bolton}. These systems also provide
causal consistency, which requires sticky availability (in each,
implemented via client-side caching). However, TA is achievable
without stickiness via an alternate algorithm: if servers wait to
reveal writes until they are present on all replicas, clients do not
need to be sticky.

We sketch a non-sticky TA algorithm and provide greater detail in our
technical report~\cite{hat-tr}. Each replica waits to reveal write $w$
from transaction $T$ until all of the writes in $T$ are present on all
respective replicas (are \textit{pending stable}). Once they are,
replicas asynchronously reveal their writes. To prevent clients from
reading writes that have been revealed on some servers but not others,
clients can attach a vector that uniquely identifies the transaction:
this can be the transaction ID and a list of keys in the
transaction. When clients attempt to read a data item from a replica,
they can send the highest ID they have observed for that item;
replicas respond with either a matching write (which may or may not be
pending stable) or a write with higher ID that is pending
stable. Clients can reset their record of IDs and keys upon
transaction commit. Servers keep two sets of versions for each data
item: the value with the highest ID that is pending stable and a set
of values that are not yet pending stable.

This non-sticky implementation is entirely masterless and neither reads
nor writes block for coordination. Moreover, there are several
possible optimizations that, for brevity, we do not describe
here. However, one key property is that we do not, in general, dictate
when writes become visible. Accordingly, it is highly
available. However, we can optimize write visibility delay by ensuring
that clients are sticky with disjoint groups of servers. Once all
transaction values are \texttt{pending} within a group of servers,
values can be made \texttt{stable} (as opposed to once they are
present on all servers). For example, in a multi-datacenter setting,
all clients and servers within a datacenter may form a sticky group.

@@

As is typical in the database literature (with few
exceptions~\cite{daudjee-session}), our models have not yet considered
interactions \textit{across} transactions, other than that there
exists some arbitrary ordering on them (via Read Uncommitted). In the
distributed systems literature, many useful \textit{safety} guarantees
span multiple, non-transactional operations. In particular,
\textit{session guarantees} are used to describe guarantees across
transactions within a given \textit{session}, ``an abstraction for the
sequence of...operations performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

@@

Berenson et al. define \textit{Lost Update} as when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider a database containing only the following
transactions:
\begin{align*}
\small\vspace{-.5em}
T_1 &: r_x(a) w_x(a+2)
\\T_2 &: w_x(2)\vspace{-.5em}
\end{align*}
If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
operation, then the database will end up with $a=3$, a state that
could not have resulted in a serial execution due to $T_2$'s
``Lost Update.''

@@

Their need to prevent Lost Update means that Consistent Read, Snapshot
Isolation, and Cursor Stability guarantees are all unavailable.
Repeatable Read (defined by Gray, Berenson et al., and Adya) and
One-Copy Serializability need to prevent both Lost Update and Write
Skew. These prevention requirements mean that these models are
inherently unavailable.

@@

To the best of our knowledge, this is the first unification of
database ACID transactions, distributed consistency, and session
guarantee models. Interestingly, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 96 possible HAT
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonomy is not exhaustive, but we believe it
lends substantial clarity into the relationships between a large
subset of the prominent ACID and distributed consistency
models. Additional read/write transaction semantics that we have
omitted should be easily classifiable based on the available
primitives and HAT-incompatible anomaly prevention we have already
discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (I-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Linearizability\linearizable, Strong
1SR\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

\end{tikzpicture}
\label{fig:hat-order}
\caption{HAT, sticky HAT (in boxes, blue), and unavailable models
  (circled, red) from Table~\protect\ref{table:hatcompared} compared
  graphically. Directed edges represent ordering by model
  strength. Models that do not share a common ancestor can be
  simultaneously achieved, and the resulting availability is that of
  the weakest available model in the combination.}
\label{fig:hatcompared}
\end{figure}

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. TA dictates that writes to
multiple keys across multiple servers are made visible to readers all
at once, while linearizability dictates that writes to a single key on
multiple servers are made visible to all readers at once---what is
different? First, in linearizable (and safe and regular) systems,
writes are made visible to clients \textit{immediately} after they
finish. With transactional atomicity, there is no recency
guarantee. Second, in linearizable systems, all clients see all writes
at the same time. With TA as defined here, clients may see writes at
different times depending on which replicas they contact. We are not
aware of an analogous model in the distributed systems
literature. Accordingly, despite apparent similarities, TA is
incomparable with and much cheaper (by availability standards) than
linearizability.

### 2013-04-28 writeup/hat-appendix.tex
some graphs and joe comments

\section{Model}

@@

@@

>>>>

\section{Adya Definitions}

@@

\section{Session Definitions}

@@

\section{Transactional Atomicity}

In our non-sticky TA algorithm, clients buffer their writes until
commit, sending the final write to each data item to each respective
server. Along with each write, clients attach a the transaction ID and
a list of all data items written to in the transaction, $L_v$. Upon
receiving a new write to item $d$, a server places the write into a
\texttt{pending} queue and notifies all other servers responsible for
a data item in $L_v$ that it has received a write to $d$ with the
respective transaction ID. Once a server receives acknowledgments with
appropriate transaction ID for all items in $L_v$, it moves the write
from \texttt{pending} to the \texttt{stable} database (which applies
the writes according to highest-ID-wins). Accordingly, all writes that
should be made atomic with a write in \texttt{stable} are guaranteed
to be present on their respective servers, either in \texttt{pending}
or \texttt{stable}. Given the possibility of internal aborts, writes
to \texttt{pending} may require 2 RTTs: one to check constraints and
one to execute the above (but these checks \textit{cannot} cause other
transactions to block, as would be the case in a two-phase
\textit{commit} protocol).

For reads, each client tracks which values it should read. It
maintains a mapping (vector) from data items to transaction IDs. Upon
transaction start, the client clears its mapping, and, when it
performs a read from a data item, it sends the server its current ID
for the item. If the client does not supply an ID, the server returns
the latest value from \texttt{stable}. If the ID is non-empty, the
server either returns the matching value from \texttt{pending} or a
value with higher ID from \texttt{pending}---along with $L_v$ for the
write. When the client receives a response, it updates its local
vector with $L_v$. Accordingly, even if transaction's writes to two
keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
and the replica for $y$'s \texttt{pending}, a client's reads from $x$
then $y$ will return a transactionally atomic pair of values because
the replica for $y$ will know to check \texttt{pending}.

### 2013-04-28 writeup/model.tex
some graphs and joe comments

Traditionally, a system provides {\em availability} if every user that
can contact a correct server eventually receives a response from that
server, even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}.\footnote{Under this
  definition, a system that requires a majority of servers to be
  online is not available.} As in a standard distributed database,
designated servers might perform operations for different data items;
a server that can handle an operation for a given data item is called
a \textit{replica} for that item.\footnote{There is a further
  distinction between a \textit{fully replicated} system, in which all
  servers are replicas for all data items, and a \textit{partially
    replicated} system, in which at least one server acts as a replica
  for only a subset of all data items. For generality, and, for
  applicability to many modern ``sharded'' or ``partitioned'' data
  storage systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we
  consider partially replicated systems.} While this standard model is
useful, there are several other cases we should consider.

@@

If a client wishes to execute a transaction that performs operations
on multiple data items, then the client must be able to contact and
receive a response from at least one replica for each data item. This
may result in ``lower availability'' than a non-transactional
requirement. Additionally, given the possibility of aborts, we need to
ensure useful forward progress: a system can trivially guarantee
clients a response by always aborting all
transactions~\cite{transaction-liveness}. However, this is an
unsatisfactory system because nothing good (transaction commit) ever
happens; we should require a \textit{liveness} property. We cannot
guarantee that every transaction will commit---transactions may choose
to abort themselves---but we need to make sure that the system will
not indefinitely abort transactions on its own volition. We call a
transaction abort due to a transaction's own choosing (e.g., as an
operation of the transaction itself or due to a would-be declared
integrity constraint violation) an \textit{internal abort} and an
abort due to system implementation an \textit{external abort}.

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally
aborts~\cite{hat-hotos}. For example, a system will violate
transactional availability if a client that can contact two
servers for each of the data items in its transaction does not
eventually commit in the absence of internal aborts.

>>>>

Traditionally, a system provides {\em availability} if every user that
can contact a correct server eventually receives a response from that
server, even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}.\footnote{Under this
  definition, systems that require a majority of servers to be online
  or that require the database to function correctly with high (but
  not perfect) probability are not available.} As in a standard
distributed database, designated servers might perform operations for
different data items; a server that can handle an operation for a
given data item is called a \textit{replica} for that
item.\footnote{There is a further distinction between a \textit{fully
    replicated} system, in which all servers are replicas for all data
  items, and a \textit{partially replicated} system, in which at least
  one server acts as a replica for only a subset of all data
  items. For generality, and, for applicability to many modern
  ``sharded'' or ``partitioned'' data storage systems~\cite{dynamo,
    pnuts, bigtable, spanner, hstore}, we consider partially
  replicated systems.} While this standard model is useful, there are
several other cases we should consider.

@@

If a client wishes to execute a transaction that performs operations
on multiple data items, then the client must be able to contact and
receive a response from at least one replica for each data item. This
may result in ``lower availability'' than a non-transactional
requirement. Additionally, given the possibility of system-initiated
aborts, we need to ensure useful forward progress: a system can
trivially guarantee clients a response by always aborting all
transactions. However, this is an unsatisfactory system because
nothing good (transaction commit) ever happens; we should require a
\textit{liveness} property~\cite{transaction-liveness}. We cannot
guarantee that every transaction will commit---transactions may choose
to abort themselves---but we need to make sure that the system will
not indefinitely abort transactions on its own volition. We call a
transaction abort due to a transaction's own choosing (e.g., as an
operation of the transaction itself or due to a would-be declared
integrity constraint violation) an \textit{internal abort} and an
abort due to system implementation an \textit{external abort}.

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits (possibly after multiple client
retries) or internally aborts~\cite{hat-hotos}. For example, a system
will violate transactional availability if a client that can contact
two servers for each of the data items in its transaction does not
eventually commit in the absence of internal aborts.
